\sys: Controllable Program Regeneration

### 2025-06-27 docs/paper/draft.tex
Transfer all to paper
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

\documentclass[a4paper,twoside,12pt]{report} %openright
\input{setup/statics.tex}
\input{setup/preamble.tex}
\input{setup/settings.tex}

@@

\pagenumbering{roman}
\input{frontmatter/frontpage.tex}
\pagecolor{white}
\newgeometry{top=2.81cm, bottom=2.75cm, outer=2.5cm, inner=3.5cm}
\pagestyle{empty}
\cleardoublepage 
\input{frontmatter/copyright.tex}
\clearpage 
\pagestyle{main}
\input{frontmatter/approval.tex}
\clearpage 
\input{frontmatter/abstract.tex}
\clearpage 
% \input{frontmatter/acknowledgements.tex}
% \cleardoublepage 
\tableofcontents
\cleardoublepage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\linenumbers
\pagenumbering{arabic}

\chapter{Introduction}
\begin{figure}[t]
  % https://docs.google.com/drawings/d/1LAmdVYjfAID24eSM-hl5PRhVg9fRMJMWYETsvvcyAuU/edit?usp=sharing
  \includegraphics[width=.9\columnwidth]{figs/kvasir_overview.pdf}
  \caption{\textbf{\sys overview}
Given a transformation query, and a source program, \sys extracts a minimal set
  of properties,
  guided by a logic engine and knowledge base.
  It then synthesizes a new program $P'$ that satisfies the query, verifying
  that wanted properties are present and unwanted properties are absent in $P'$.
}
  \label{fig:overview}
\end{figure}

@@

% TODO: Add chapter references
\heading{Contributions}
The paper's contributions are:
\begin{itemize}
  \item A property-aware transformation framework, combining logic-based reasoning, and LLM-assisted program synthesis~(\cref{sec:design});
 \item A declarative interface and accompanying DSL for program transformation, allowing users to express semantic goals as logical queries over properties of the regenerated program~(\cref{sec:dsl});
 \item A verification architecture that checks synthesized programs against the specified properties, providing feedback for refinement~(\cref{sec:verification});
 \item An empirical evaluation demonstrating \sys's effectiveness and scalability across transformation scenarios~(\cref{sec:evaluation}).
\end{itemize}

@@

\chapter{\sys Implementation}

@@

\section{Example}
\label{sec:example}
This chapter shows the application of \sys
to three distinct program transformation tasks, 
targeting security, language translation, compatibility,
idiomaticity, and modularity.

@@

\section{Overall Design}
\label{sec:design}

@@

\section{Property Language and Extraction}
\label{sec:dsl}

@@

\section{Synthesis Plan}
\label{sec:synthesis}

@@

\section{Property Verification and Feedback Loop}
\label{sec:verification}

@@

\chapter{Evaluation}
\label{sec:evaluation}

@@

\section{Benchmarks \& Methodology}

@@

\begin{table}[h]
\centering
  \caption{\textbf{Benchmark summary}. 
  Benchmark programs come from online repositories, public archives, security research datasets, and often referenced examples in programming culture.
The N column shows the number of programs/functions from each benchmark.
}
\begin{tabular}{llrll}
\toprule
Benchmark                          & Description         & N  & Task                                       & Source \\
\midrule
Rosetta Code                       & Simple tasks   & 16 & Translation                       & \cite{rosettacode} \\
Short utilities                    & npm/PyPi utilities & 13 & Parity & \cite{regbench2025} \\
SSCA                               & Sneaky SCAs         & 3  & Attack elimination                 & \cite{ev:eurosec:2022, ohm2020backstabber,copeland2019frightening} \\
IOCCC                              & Obfuscated C  & 2  & De-obfuscation                             & \cite{ioccc} \\
Microbenchmarks                    &                     & 3  &                                            & \\
\hspace{.5em} \ttt{flatmap-stream} & Flat-map utility &    & Purity \& translation & \cite{es1}  \\
\hspace{.5em} \ttt{Q\_rsqrt}       & Reverse sqrt       &    & Idiomatic rewrite                          & \cite{fast_inv_sqrt}  \\
\hspace{.5em} \textsf{rigidDB}     & Music database      &    & Modularization                             & \cite{codewithsadeemusicplayer} \\
\midrule
Total                              &                     & 37 &                                            & \\
\bottomrule
\end{tabular}
\label{tab:benchmarks}
\end{table}

\section{Q1: Correctness and Regeneration Quality}

@@

\begin{table}[h]
  \centering
  \caption{Correctness: Number of tasks passing all tests (when present) and manual inspection.}
  \begin{tabular}{lc}
    \toprule
    Benchmark & Correctness \\ 
    \midrule
    Rosetta Code & 16/16  \\
    Short Utilities & 13/10 \\
    SSCA & 3/3  \\
    IOCCC & 1/2 \\
    Microbenchmarks & 2.5/3  \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\section{Q2: Naive LLM vs. \sys}

@@

\section{Q3: Planning Efficiency \& Precision}

@@

\chapter{Related Work}

@@

\chapter{Discussion and Limitations}
\label{sec:discussion}

@@

\chapter{Conclusion}
This paper introduced \sys, a system for property-guided program regeneration
that bridges the flexibility of large language models with the precision of
logic-driven planning.
By allowing developers to declaratively specify
transformation goals as logical constraints, \sys elevates LLM-assisted program
transformations to a controllable, interpretable, and iterative synthesis-verification
process.
Evaluation demonstrates that \sys produces high-quality regenerations
across diverse tasks, outperforming naive prompting baselines in correctness,
controllability, and resilience to adversarial or obfuscated inputs.

@@

\heading{Ethical and legal considerations}
While this work has applied \sys exclusively to open-source software, its
capabilities raise ethical and legal questions. Regenerating
proprietary software, de-obfuscating binaries, or reimplementing licensed
systems into equivalent ones with more permissive licensing 
could challenge existing norms and
regulations. 
It is important to consider
safeguards and responsible disclosure practices towards ethical program regeneration.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\printbibliography[heading=bibintoc,title={Bibliography}]
\cleardoublepage 
\appendix
\input{backmatter/appendix.tex}
\cleartoleftpage
\input{backmatter/backpage.tex}

>>>>

\documentclass[noacm,sigplan,review]{acmart}

% \acmSubmissionID{PAPER ID}
\renewcommand\footnotetextcopyrightpermission[1]{}
\settopmatter{printfolios=false,printacmref=false}

\usepackage{setspace}
\usepackage{enumerate}
\usepackage{algorithm2e}
\usepackage{algpseudocode}
\usepackage{graphics}
\usepackage{xparse} 
\usepackage{xspace}
\usepackage{multirow}
\usepackage{csvsimple}
\usepackage{balance}
\usepackage{minted}
\usepackage{xcolor}
\usepackage{nicefrac}
\usepackage{siunitx}
\usepackage{array,framed}
\usepackage{booktabs}
\usepackage{color}
\usepackage{soul}
\usepackage{float}
\usepackage{epsfig}
\usepackage{wrapfig}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{adjustbox}
\usepackage{csquotes}
\usepackage{cleveref}
\usepackage{dirtytalk}
\usepackage{textgreek}
\usepackage{oplotsymbl}
\usepackage{listings}
\usepackage{flushend}

\def\eg{{\em e.g.}, }
\def\ie{{\em i.e.}, }
\def\etc{{\em etc.}\xspace}
\def\vs{{\em vs.}\xspace}

\def\gptmodel{{GPT-4o}\xspace}

\newcommand{\todo}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\sys}{{\scshape Kv{\textalpha}sir}\xspace}
\newcommand{\rf}[1]{\ref{#1}}
\newcommand{\sx}[1]{(\S\ref{#1})}
\newcommand{\cf}[1]{(\emph{Cf}.\S\ref{#1})}
\newcommand{\se}[1]{\S\ref{#1}}
\newcommand{\fg}[1]{Fig.~\ref{#1}}
\newcommand{\heading}[1]{\vspace{2pt}\noindent\textbf{\emph{#1}}:\enspace}
\newcommand{\ttt}[1]{\texttt{#1}\xspace}
\newcommand{\xxx}{\colorbox{red!30}{xxx}\xspace}
\newcommand{\promptbox}[1]{\hspace*{-0.0135\columnwidth}\fbox{\parbox[t]{0.95\columnwidth}{\tt{#1}}}}

\crefformat{section}{\S#2#1#3}
\crefmultiformat{section}{\S#2#1#3}{--\S#2#1#3}{, \S#2#1#3}{ and \S#2#1#3}

@@

\title{\sys: Controllable Program Regeneration}
\author{Evangelos Lamprou}

% Use cases for something like this include:
% - program repair
% - turning insecure code into secure code by removing side-effects
% - transforming a program in language A to language B
% - turning a program into a more idiomatic version of the same language
% - having a program use a different API
% - have a program be more amendable to parallelization
% - have a program be more amendable to further analysis or transformation

\begin{abstract}
Modern software systems are constantly evolving—requiring developers to
  refactor legacy code, adapt to changing APIs, translate across languages, and
  patch security vulnerabilities. These transformations are often brittle,
  manual, and error-prone. Large language models (LLMs) offer new opportunities
  for automating such tasks, but their outputs are shaped by surface cues, lack
  semantic guarantees, and require brittle prompting strategies. This paper
  presents \sys, a system for controllable program regeneration that combines
  logic-based planning with LLM-guided synthesis.
Kvasir operates in four stages: (1) a symbolic planner constructs a
transformation plan using logic rules and a domain-aware knowledge base; (2)
property extractors analyze the input program to gather relevant semantic
information; (3) a language model synthesizes candidate programs guided by
a structured prompt assembled from extracted properties; and (4) a verification
and feedback loop checks whether the output meets the desired properties and
iteratively refines the generation process through additional feedback.
\sys is evaluated on a diverse set of program transformation tasks
including security hardening, language translation, idiomatic rewriting, de-obfuscation,
and modular refactoring.
Target programs are sourced from high-profile supply-chain attacks, 
online repositories, programming competitions, and real-world software projects.
\sys produces high-quality outputs that satisfy goals in all but two cases,
even on instances where a state-of-the-art LLM fails to do so.
\end{abstract}
\maketitle

\section{Introduction}
\begin{figure}[t]
  % https://docs.google.com/drawings/d/1LAmdVYjfAID24eSM-hl5PRhVg9fRMJMWYETsvvcyAuU/edit?usp=sharing
  \includegraphics[width=.9\columnwidth]{figs/kvasir_overview.pdf}
  \caption{\textbf{\sys overview}
Given a transformation query, and a source program, \sys extracts a minimal set
  of properties,
  guided by a logic engine and knowledge base.
  It then synthesizes a new program $P'$ that satisfies the query, verifying
  that wanted properties are present and unwanted properties are absent in $P'$.
}
  \label{fig:overview}
\end{figure}

@@

% TODO: Add section references
\heading{Contributions}
The paper's contributions are:
\begin{itemize}
  \item A property-aware transformation framework, combining logic-based reasoning, and LLM-assisted program synthesis~(\cref{sec:design});
 \item A declarative interface and accompanying DSL for program transformation, allowing users to express semantic goals as logical queries over properties of the regenerated program~(\cref{sec:dsl});
 \item A verification architecture that checks synthesized programs against the specified properties, providing feedback for refinement~(\cref{sec:verification});
 \item An empirical evaluation demonstrating \sys's effectiveness and scalability across transformation scenarios~(\cref{sec:evaluation}).
\end{itemize}

@@

\section{\sys Implementation}

@@

\subsection{Example}
\label{sec:example}
This section shows the application of \sys
to three distinct program transformation tasks, 
targeting security, language translation, compatibility,
idiomaticity, and modularity.

@@

\subsection{Overall Design}
\label{sec:design}

@@

\subsection{Property Language and Extraction}
\label{sec:dsl}

@@

\subsection{Synthesis Plan}
\label{sec:synthesis}

@@

\subsection{Property Verification and Feedback Loop}
\label{sec:verification}

@@

\section{Evaluation}
\label{sec:evaluation}

@@

\subsection{Benchmarks \& Methodology}

@@

\begin{table*}[h]
\centering
  \caption{\textbf{Benchmark summary}. 
  Benchmark programs come from online repositories, public archives, security research datasets, and often referenced examples in programming culture.
The N column shows the number of programs/functions from each benchmark.
}
  \begin{tabular*}{\textwidth}{llrrll}
\toprule
Benchmark                          & Description        & N  & LoC  & Task                   & Source \\
\midrule
Rosetta Code                       & Simple tasks       & 16 &      & Translation            & \cite{rosettacode} \\
Short utilities                    & npm/PyPi utilities & 13 &      & Parity                 & \cite{regbench2025} \\
SSCA                               & Sneaky supply-chain attacks        & 3  &      & Attack elimination     & \cite{ev:eurosec:2022, ohm2020backstabber,copeland2019frightening} \\
IOCCC                              & Obfuscated C code       & 2  &      & De-obfuscation         & \cite{ioccc} \\
Microbenchmarks                    &                    & 3  &      &                        & \\
\hspace{.5em} \ttt{flatmap-stream} & Flat-map utility   &    &      & Purity and translation & \cite{es1}  \\
\hspace{.5em} \ttt{Q\_rsqrt}       & Inverse square root function       &    &      & Idiomatic rewrite      & \cite{fast_inv_sqrt}  \\
\hspace{.5em} \textsf{rigidDB}     & Fetch artist songs from a music database     &    &      & Modularization         & \cite{codewithsadeemusicplayer} \\
\midrule
Total                              &                    & 37 &      &                        & \\
\bottomrule
\end{tabular*}
\label{tab:benchmarks}
\end{table*}

\subsection{Q1: Correctness and Regeneration Quality}

@@

\begin{table}[h]
  \centering
  \caption{\textbf{Correctness result summary.} Number of tasks passing all tests (when present) and manual inspection.}
  \begin{tabular}{lc}
    \toprule
    Benchmark & Correctness \\ 
    \midrule
    Rosetta Code & 16/16  \\
    Short Utilities & 13/10 \\
    SSCA & 3/3  \\
    IOCCC & 1/2 \\
    Microbenchmarks & 2.5/3  \\
    \bottomrule
  \end{tabular}
\end{table}

@@

\subsection{Q2: Naive LLM vs. \sys}

@@

\subsection{Q3: Planning Efficiency \& Precision}

@@

\section{Related Work}

@@

\section{Discussion and Limitations}
\label{sec:discussion}

@@

\section{Conclusion}
This paper introduced \sys, a system for property-guided program regeneration
that bridges the flexibility of large language models with the precision of
logic-driven planning.
By allowing developers to declaratively specify
transformation goals as logical constraints, \sys elevates LLM-assisted program
transformations to a controllable, interpretable, and iterative synthesis-verification
process.
Evaluation demonstrates that \sys produces high-quality regenerations
across diverse tasks, outperforming naive prompting baselines in correctness,
controllability, and resilience to adversarial or obfuscated inputs.

@@

\heading{Ethical and legal considerations}
While this work has applied \sys exclusively to open-source software, its
capabilities raise ethical and legal questions. Regenerating
proprietary software, de-obfuscating binaries, or reimplementing licensed
systems into equivalent ones with more permissive licensing 
could challenge existing norms and
regulations. 
It is important to consider
safeguards and responsible disclosure practices towards ethical program regeneration.

\bibliographystyle{ACM-Reference-Format}
\bibliography{bib.bib}

\appendix
